Univalent Universes and Completness of Reversible Programs: \\ Featherweight HoTT

### 2017-03-06 upi.tex
cleaning up definitions

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (-3,0) ellipse (1.5cm and 3cm);
  \draw (3,2) ellipse (0.5cm and 1cm);
  \draw (3,-2) ellipse (0.5cm and 1cm);
  \node[blue,ultra thick,above] at (-3,3) {$A$};
  \node[blue,ultra thick,above] at (3,3) {$P(x)$ (fiber over $x$)};
  \node[blue,ultra thick,below] at (3,-3) {$P(y)$ (fiber over $y$)};
  \draw[fill] (-3,1.5) circle [radius=0.025];
  \draw[fill] (-3,-1.5) circle [radius=0.025];
  \draw[left,cyan,thick] (-3,1.5) -- (-3,-1.5);
  \node[above] at (-3,1.5) {$x$};
  \node[below] at (-3,-1.5) {$y$};
  \draw[fill] (3,1.8) circle [radius=0.025];
  \draw[fill] (3,-1.8) circle [radius=0.025];
  \node[above] at (3,1.8) {$u$};
  \node[below] at (3,-1.8) {$p_*(u)$};
  \node[left,cyan] at (-3,0) {$p$};
  \draw[->,red,dashed,ultra thick] (-3,1.5) to [out=45, in=135] (2.4,2.5);
  \draw[->,red,dashed,ultra thick] (-3,-1.5) to [out=-45, in=-135] (2.4,-2.5);
\end{tikzpicture}
\end{center}

\paragraph*{Univalent Type Family.} A type family $P : A \to \mathcal{U}$ is
\emph{univalent} if the following function is an equivalence:
\[\begin{array}{rcl}
f & : & (x \equiv y) \to (P(x) \simeq P(y)) \\
\end{array}\]
where $x,y:A$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Univalence}

Let $X : \Typ$. We have:
\[
\textit{transport-equiv} : \Pi_{P : X \to \Typ}~ \Pi_{x,y:X}~
    x \equiv y  \to P x \simeq P y
\]
For a fibration with base $X$, define:
\[\begin{array}{l}
\textit{is-univalent-fibration} : (X \to \Typ) \to \Typ \\
\textit{is-univalent-fibration}~P =
    \Pi_{x,y:X}~\textit{is-equiv} (\textit{transport-equiv}~P~x~y)
\end{array}\]

\noindent\paragraph*{Examples}
\begin{itemize}
\item Take $X = \Typ$ and $P = \mathit{Id}_{\Typ}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\Typ}~x \equiv y \to
  x \simeq y$ which is the usual definition of univalence.
\item Take $X = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
  statement $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
\item Take $X = \ot$ and $P = \lambda \_. \zt$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
\item Take $X = \ot$. For no $P$ other than the above two instances will
  we have a univalent fibration.
\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \_. \ot$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\ot\simeq\ot)$ which is false.
\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\zt\simeq\ot)$ which is true. That is the only choice of $P$
  that works (modulo symmetry).
\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\ot$ can never
  give a univalent fibration.
\end{itemize}

\begin{theorem}
  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F
  \|$. We will refer to this type as $\{F\}$.
\end{theorem}

\noindent In particular $\{\bt\}$, i.e.,
$\Sigma_{Y : \Typ} \| Y \equiv \bt \|$ is the base of a univalent
fibration. In some sense this is the simplest univalent universe, our
featherweight HoTT that we aim to characterize as a reversible programming
language based on type isomorphisms.

>>>>

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (-3,0) ellipse (1.5cm and 3cm);
  \draw (3,2) ellipse (0.5cm and 1cm);
  \draw (3,-2) ellipse (0.5cm and 1cm);
  \node[blue,ultra thick,above] at (-3,3) {$A$};
  \node[blue,ultra thick,above] at (3,3) {$P(x)$ (fiber over $x$)};
  \node[blue,ultra thick,below] at (3,-3) {$P(y)$ (fiber over $y$)};
  \draw[fill] (-3,1.5) circle [radius=0.025];
  \draw[fill] (-3,-1.5) circle [radius=0.025];
  \draw[left,cyan,thick] (-3,1.5) -- (-3,-1.5);
  \node[above] at (-3,1.5) {$x$};
  \node[below] at (-3,-1.5) {$y$};
  \draw[fill] (3,1.5) circle [radius=0.025];
  \draw[fill] (3,2.1) circle [radius=0.025];
  \draw[fill] (3,-1.8) circle [radius=0.025];
  \node[above] at (3,1.5) {$u$};
  \node[above] at (3,2.1) {$v$};
  \node[below] at (3,-1.8) {$p_*(u)$};
  \node[left,cyan] at (-3,0) {$p$};
  \draw[->,red,dashed,ultra thick] (-3,1.5) to [out=45, in=135] (2.4,2.5);
  \draw[->,red,dashed,ultra thick] (-3,-1.5) to [out=-45, in=-135] (2.4,-2.5);
\end{tikzpicture}
\end{center}

It is worth repeating Remark 2.7.1 from the book. In the figure above, we could
have $(x,u) \equiv (x,v)$ without having $u \equiv v$. The only thing we can
conclude is that there is a (possibly non-trivial) path path $q : x \equiv x$
such that $q_*(u) \equiv v$.

\paragraph*{Univalent Type Family.}
Given standard constructions in the book, we have a function:
\[\begin{array}{rcl}
\textit {transport-equiv} &:& \Pi_{P : A \to \mathcal{U}}~ \Pi_{x,y:A}~
    x \equiv y  \to P x \simeq P y \\
\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoequiv}(\mathit{ap}_{P}(p))
\end{array}\]
with the standard definitions for
$\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y))$ and
$\mathit{idtoequiv} : (A \equiv B) \to (A \simeq B)$. We say a type family
$P : A \to \mathcal{U}$ is \emph{univalent} if for all $x,y:A$
the function $\textit{transport-equiv}~P~x~y$ is an equivalence. Formally:
\[\begin{array}{l}
\textit{is-univalent-fibration} : (A \to \mathcal{U}) \to \mathcal{U} \\
\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{is-equiv}~(\textit{transport-equiv}~P~x~y)
\end{array}\]

\noindent\paragraph*{Examples}
\begin{itemize}
\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\mathcal{U}}~x \equiv y \to
  x \simeq y$ which is the usual definition of univalence.

\newpage

\item Take $X = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
  statement $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
\item Take $X = \ot$ and $P = \lambda \_. \zt$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
\item Take $X = \ot$. For no $P$ other than the above two instances will
  we have a univalent fibration.
\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \_. \ot$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\ot\simeq\ot)$ which is false.
\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\zt\simeq\ot)$ which is true. That is the only choice of $P$
  that works (modulo symmetry).
\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\ot$ can never
  give a univalent fibration.
\end{itemize}

\begin{theorem}
  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
  if there exists an $F$ such that $X \simeq \Sigma_{Y : \mathcal{U}} \| Y \equiv F
  \|$. We will refer to this type as $\{F\}$.
\end{theorem}

\noindent In particular $\{\bt\}$, i.e.,
$\Sigma_{Y : \mathcal{U}} \| Y \equiv \bt \|$ is the base of a univalent
fibration. In some sense this is the simplest univalent universe, our
featherweight HoTT that we aim to characterize as a reversible programming
language based on type isomorphisms.
